\documentclass[twocolumn,superscriptaddress,amsmath,amssymb,aps,floatfix]{revtex4-2}
\usepackage{graphicx,tikz} 
\usepackage[mode=buildmissing]{standalone} 
\usepackage{dcolumn}
\usepackage{bm}
\usepackage{physics}
\usepackage{siunitx}
\usepackage{soul}
\usepackage{comment}
\usepackage[bb=boondox]{mathalfa}
\usetikzlibrary{external}
\tikzexternalize[prefix=figures/]
\DeclareMathOperator{\sinc}{sinc}
\newcommand{\may}[1]{\textcolor{orange}{#1}}
\newcommand{\ale}[1]{{\color{blue}{\bf Ale}: #1}}
\newcommand{\zd}[1]{{\color{purple}{\bf ZD}: #1}}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tcolorbox}
\usepackage{tabularx}
\usepackage{array}
\usepackage{colortbl}
\tcbuselibrary{skins}
\newcolumntype{Y}{>{\raggedleft\arraybackslash}X}

\tcbset{tab1/.style={fonttitle=\bfseries\large,fontupper=\normalsize\sffamily,
colback=yellow!10!white,colframe=red!75!black,colbacktitle=Salmon!40!white,
coltitle=black,center title,freelance,frame code={
\foreach \n in {north east,north west,south east,south west}
{\path [fill=red!75!black] (interior.\n) circle (3mm); };},}}

\tcbset{tab2/.style={enhanced,fonttitle=\bfseries,fontupper=\normalsize\sffamily,
colback=yellow!10!white,colframe=red!50!black,colbacktitle=Salmon!40!white,
coltitle=black,center title}}

\begin{document}
\title{Proposal for a Nuclear Talent course at the ECT* in 2025:\\Quantum Computing for Nuclear
Physics}


\author{Alexei Bazavov}

\affiliation{Department of Computational Mathematics, Science and
Engineering and Department of Physics and Astronomy, Michigan State
University, East Lansing, MI 48824, USA}

\author{Zohreh Davoudi}
\affiliation{Department of Physics, University of Maryland, College Park,
MD 20742, USA}
\affiliation{Center for Quantum Information and Computer Science (QuICS), University of Maryland, College Park, MD 20742, USA}

\author{Morten Hjorth-Jensen}
\affiliation{Department of Physics and Center for Computing in Science Education, University of Oslo, N-0316 Oslo, Norway}
\affiliation{Facility for Rare Isotope Beams and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}

\author{Ryan LaRose}
\affiliation{Department of
Computational Mathematics, Science and Engineering, Michigan State
University, East Lansing, MI 48824, USA}

\author{Dean Lee}
\affiliation{Facility for Rare Isotope Beams and Department of Physics and
Astronomy, Michigan State University, East Lansing, MI 48824, USA}

\author{Alessandro Roggero} 
\affiliation{Department of Physics, University of Trento, Povo, 38123 Trento, Italy}




    
\maketitle
    
    


    \section{Proposal for a Nuclear Talent course at the ECT* in
2025}\label{proposal-for-a-nuclear-talent-course-at-the-ect-in-2025}

We would like to propose a three-week Nuclear Talent course on theory
for exploring quantum computing and quantum information science applied
to nuclear physics at the ECT* for the summer of 2025. We believe such a
course has a strong potential to attract many students, theorists and
experimentalists alike.

Below we give a short motivation for the proposed course and the
rationale behind the Nuclear Talent initiative. Thereafter we detail our
course plans with learning outcomes, objectives and teaching philosophy,
as well as various organizational and practical matters.

The teaching teams consists of quantum computing theorists and nuclear
theorists, with expertise ranging from many-body methods to quantum
field theories. Several of us have developed courses on quantum
computing and/or taught similar courses, in addition to our ongoing
research on quantum computing. This spans from quantum engineering to
developments of new algorithms and error correction studies. We believe
such a mix is important as it gives the students a better understanding
on how quantum computing can be applied to nuclear physics problems, and
what are the limitations and possibilities in understanding and
interpreting the various algorithms on present and planned quantum
technologies.

    \subsection{Motivation}\label{motivation}

For nuclear theorists, the overarching challenge is to develop a
comprehensive description of nuclei and their reactions, grounded in the
fundamental interactions between the constituent nucleons with
quantifiable uncertainties to maximize predictive power. As experimental
frontiers have shifted to the study of rare isotopes, the predictive
power of successful phenomenological approaches like the shell model and
density functional theory is challenged by the scarcity of nearby
experimental data to constrain model parameters. Therefore, it is
expected that few-body and many-body methods will play an increasingly
prominent role to help improve the predictive power of such ``data
driven'' methods as experiment moves deeper into largely unexplored
regions of the nuclear chart.

To understand why nuclear matter is stable, and thereby shed light on
the limits of nuclear stability, is one of the overarching aims and
intellectual challenges of basic research in nuclear physics. To relate
the stability of matter to the underlying fundamental forces and
particles of nature as manifested in nuclear matter, is central to
present and planned rare isotope facilities. From a theoretical
standpoint, this involves understanding how the basic building blocks of
Nature interact and conspire to build up atomic nuclei as we know them,
with the aim to understand what makes visible matter stable. The
theoretical efforts span from methods like lattice quantum
chromodynamics, via effective field theories to many-body theories
applied to atomic nuclei and infinite nuclear matter. All these methods
rely on theoretical approximations whose applicabilities are often
limited by the dimensionality of the specific problem being studied. In
recent years, there has been considerable progress in 
%our theoretical understanding of 
developing quantum-computing algorithms
%from quantum information theories 
applied to quantum many-body systems, with the hope to circumvent many of the 
%traditional dimensionality 
classically intractable problems.

This proposal for a Nuclear Talent school aims at bringing together the
efforts of nuclear many-body theorists, quantum information theorists,
and mathematicians in order to present and discuss algorithms for
studying nuclear systems using recent progress in quantum information
theory.

    \subsection{Introduction to the Talent
Courses}\label{introduction-to-the-talent-courses}

The TALENT initiative, \href{http://www.nucleartalent.org}{Training in
Advanced Low Energy Nuclear Theory}, aims at providing an advanced and
comprehensive training to graduate students and young researchers in
low-energy nuclear theory. The initiative is a multinational network
of several European and Northern American institutions and aims at
developing a broad curriculum that will provide %the platform for a
the necessary training in cutting-edge theory for understanding nuclei and nuclear reactions.
These objectives will be met by offering a series of lectures,
%commissioned from 
delivered by 
%experienced instructors 
experts in nuclear theory and experiment.
The educational material generated under this program will be collected
in the form of WEB-based courses, textbooks, and a variety of modern
educational resources. No such all-encompassing material is available at
present; its development will allow dispersed university groups to
profit from the best expertise available worldwide.

The Nuclear Talent initiative has organized and run several advanced
courses since the summer of 2012. Several of these courses have been run
and organized (in a very successful way) at the premises of the
ECT\emph{. We hope thus, if this course gets approved by the board of
directors of the ECT}, that we can continue this successful and very
fruitful collaboration.

    \section{Aims and Learning
Outcomes}\label{aims-and-learning-outcomes}

    \subsubsection{Format:}\label{format}

We propose approximately forty-five hours of lectures over three weeks
and a comparable amount of 
%practical computer and exercise sessions,
supervised practices and tutorials,
including 
%the setting of individual 
with the aid of well-thought-of problem sets and 
%the organization of
various individual projects.

The mornings will consist of lectures and the afternoons will be devoted
to exercises meant to shed light on the exposed theory, the
computational projects, and individual student projects. These components
will be coordinated to foster student engagement, maximize learning, and
create lasting value for the students. For the benefit of the TALENT
series and of the community, material (courses, slides, problems and
solutions, reports on students' projects) will be made publicly
available using version control software like \emph{git} and posted
electronically on \href{https://github.com}{github}.

As with previous TALENT courses, we envision the following features for
the afternoon sessions: * We will use both individual and group work to
carry out tasks that are very specific in technical instructions, but
leave freedom for creativity.

\begin{itemize}
\item
  Groups will be carefully put together to maximize diversity of
  backgrounds.
\item
  Results will be presented in a conference-like setting to create
  accountability.
\item
  We will organize events where individuals and groups exchange their
  experiences, difficulties, and successes to foster interaction.
\item
  During the school, on-line and lecture-based training tailored to
  technical issues will be provided. Students will learn to use and
  interpret the results of computer-based and hands-on calculations of
  quantum computing algorithms. The lectures will be aligned with the
  practical computational projects and exercises and the lecturers will
  be available to help students and work with them during the exercise
  sessions.
\item
  These interactions will raise topics not originally envisioned for the
  course but which are recognized to be valuable for the students. There
  will be flexibility to organize mini-lectures and discussion sessions
  on an ad-hoc basis in such cases.
\item
  Each group of students will maintain an online logbook of their
  activities and results.
\item
  Training modules, codes, lectures, practical exercise instructions,
  online logbooks, instructions and information created by participants
  will be merged into a comprehensive website that will be available to
  the community and the public for self-guided training or for use in
  various educational settings (for example, a graduate course at a
  university could assign some of the projects as homework or an extra
  credit project, etc).
\end{itemize}

    \subsubsection{Objectives and learning
outcomes:}\label{objectives-and-learning-outcomes}

At the end of the course the students should have a basic understanding
of (these will be updated). More material will be added here.

    \subsection{Course Content and detailed plan}
%(\textbf{tentative, not
%final})}{Course Content and detailed plan (tentative, not final)}}\label{course-content-and-detailed-plan-tentative-not-final}


    \subsection{Teaching}\label{teaching}

The course will be taught as an intensive course of duration of three
weeks, with a total time of 45 h of lectures, 45 h of exercises and a
final assignment of 2 weeks of work. The total load will be
approximately 160-170 hours, corresponding to \textbf{7 ECTS} in Europe.
The final assignment will be graded with marks A, B, C, D, E and failed
for Master students and passed/not passed for PhD students. A course
certificate will be issued for students requiring it from the University
of Trento.

The organization of a typical course day is as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  9am-12pm: Lectures, project relevant information and directed
  exercises
\item
  12pm-2pm: Lunch
\item
  2pm-6pm: Computational projects, exercises and hands-on sessions
\item
  6pm-7pm: Wrap-up of the day and eventual student presentations
\end{enumerate}

If approved by the ECT* board of directors, our preferred time slot
would be from the second half of June till the second half of July.

    \subsection{Instructors and organizers}\label{instructors-and-organizers}

The organizers and instructors are \zd{It would be good to mention what each instructor's area of expertise is. For me, you can add: Davoudi is an expert in lattice QCD for nuclear physics. She further specializes in quantum simulation and computing for quantum field theories, including lattice gauge theories and effective field theories of nuclear physics.}


\begin{enumerate}
\item 
\href{https://directory.natsci.msu.edu/Directory/Profiles/Person/101033}{Alexei
Bazavov} at Department of Computational Mathematics, Science and
Engineering and Department of Physics and Astronomy, Michigan State
University, East Lansing, MI 48824, USA
\item
  \href{https://umdphysics.umd.edu/people/faculty/current/item/927-davoudi.html}{Zohreh
  Davoudi}, Department of Physics and Center for Quantum Information and Computer Science (QuICS), University of Maryland, College Park,
  MD 20742, USA
\item
  \href{http://mhjgit.github.io/info/doc/web/}{Morten Hjorth-Jensen} at
  Facility for Rare Isotope Beams and Department of Physics and
  Astronomy, Michigan State University, East Lansing, MI 48824, USA \&
  Department of Physics, University of Oslo, N-0316 Oslo, Norway
\item
  \href{https://frib.msu.edu/for-students/faculty/lee-profile}{Dean Lee}
  at Facility for Rare Isotope Beams and Department of Physics and
  Astronomy, Michigan State University, East Lansing, MI 48824, USA
\item
  \href{https://www.ryanlarose.com/}{Ryan LaRose} at at Department of
  Computational Mathematics, Science and Engineering, Michigan State
  University, East Lansing, MI 48824, USA
\item
  \href{https://webapps.unitn.it/du/en/Persona/PER0016084/Didattica}{Alessandro
  Roggero}, Department of Physics, University of Trento, Povo, 38123
  Trento, Italy
\item
  An eventual Post-doctoral fellow and an advanced graduate student as
  teaching assistants.
\end{enumerate}

Morten Hjorth-Jensen Alessandro Roggero will also function as student
advisors and coordinators.

    \subsection{Audience and
Prerequisites}\label{audience-and-prerequisites}

Students and post-doctoral fellows interested in quantum computing
applied to nuclear physics problems, from nuclear structure to quantum
field theories. The material will be of interest, and accessible, to
both theorists and experimentalists, and will include learning the
practical use of quantum computing software in order to interpret and
study nuclear systems.

The students are expected to have operating programming skills in
compiled programming languages like Fortran or C++ or preferentially in
an interpreted language like Python and knowledge of quantum mechanics
at an intermediate level.

    \subsection{Admission}\label{admission}

The target group is Master of Science students, PhD students and early
post-doctoral fellows. Also senior staff can attend but they have to be
self-supported. The maximum number of students is 20-30, of which
hopefully at most 15-20 can receive full local support.

The process of selections of the students will be managed in agreement
with the ECT*.

    \subsection{Preliminary budget}\label{preliminary-budget}

We expect to accept between 20-30 students. Local students from the
University of Trento are fully self-supported. If approved, we would
very much appreciate if the ECT* can sponsor 15-20 of the selected
students with local expenses, that is lodging and meals during weekdays.
Any additional funds for sponsoring further students is highly
appreciated.

All travel expenses will be covered by the respective home institute.
Instructors are self-supported. We plan to raise additional funds to cover
local support for additional students and the expenses of the instructors.

There is no participation fee. Administrative support from the ECT* in
organizing the course and setting up the application procedure is
essential for a smooth (as always) outcome. The administrative
experience of the staff at the ECT* has been unique and essential in
running successfully our previous Talent courses. We would thus highly
appreciate it if these services are provided if the proposal is
approved.


\begin{tcolorbox}[tab2,tabularx={X||Y|Y|Y|Y||Y}]
Group & One     & Two     & Three    & Four     & Sum      \\\hline\hline
Red   & 1000.00 & 2000.00 &  3000.00 &  4000.00 & 10000.00 \\\hline
Green & 2000.00 & 3000.00 &  4000.00 &  5000.00 & 14000.00 \\\hline
Blue  & 3000.00 & 4000.00 &  5000.00 &  6000.00 & 18000.00 \\\hline\hline
Sum   & 6000.00 & 9000.00 & 12000.00 & 15000.00 & 42000.00
\end{tcolorbox}

\begin{tcolorbox}[tab2,tabularx={X||Y|Y|Y|Y||Y},title=My table,boxrule=0.5pt]
Group & One     & Two     & Three    & Four     & Sum      \\\hline\hline
Red   & 1000.00 & 2000.00 &  3000.00 &  4000.00 & 10000.00 \\
Green & 2000.00 & 3000.00 &  4000.00 &  5000.00 & 14000.00 \\
Blue  & 3000.00 & 4000.00 &  5000.00 &  6000.00 & 18000.00 \\\hline\hline
Sum   & 6000.00 & 9000.00 & 12000.00 & 15000.00 & 42000.00
\end{tcolorbox}

\begin{tcolorbox}[tab1,tabularx={X||YYYY||Y}]
Group & One     & Two     & Three    & Four     & Sum      \\\hline\hline
Red   & 1000.00 & 2000.00 &  3000.00 &  4000.00 & 10000.00 \\
Green & 2000.00 & 3000.00 &  4000.00 &  5000.00 & 14000.00 \\
Blue  & 3000.00 & 4000.00 &  5000.00 &  6000.00 & 18000.00 \\\hline\hline
Sum   & 6000.00 & 9000.00 & 12000.00 & 15000.00 & 42000.00
\end{tcolorbox}



\end{document}


Week 1: Quantum computing preliminaries

[Day 1, Lect 1] Review of QM and linear algebra
[Day 1, Lect 2] Qubits, gates, and circuits
[Day 2, Lect 1] Introduction to quantum algorithms
[Day 2, Lect 2] Quantum fourier transform and quantum phase estimation
[Day 3, Lect 1] Quantum simulation: product formula algorithms
[Day 3, Lect 2] Quantum simulation: non-product formula algorithms
[Day 4, Lect 1] Variational quantum algorithms
[Day 4, Lect 2] Quantum measurements and state tomography
[Day 5, Lect 1] Noise mitigation and NISQ computing
[Day 5, Lect 2] Quantum error correction and fault tolerance

Week 1: Quantum computing for nuclear/neutrino many-body physics

[Day 1, Lect 1] Encoding (relativistic and nonrelativistic) fermions and bosons on quantum computers I
[Day 1, Lect 2] Encoding (relativistic and nonrelativistic) fermions and bosons on quantum computers I
[Day 2, Lect 1] Quantum simulation of pionless EFT
[Day 2, Lect 2] Quantum simulation of pionfull EFT
[Day 3, Lect 1] Quantum simulation of phenomenological nuclear models
[Day 3, Lect 2] Quantum state preparation (ground, excited, and thermal states) I
[Day 4, Lect 1] Quantum state preparation (ground, excited, and thermal states) II
[Day 4, Lect 2] Physics target: nuclear response functions
[Day 5, Lect 1] Physics target: neutrino dynamics in dense environments
[Day 5, Lect 2] Entanglement in nuclear many-body systems

Week 1: Quantum computing for quantum (gauge) field theories

[Day 1, Lect 1] Quantum simulation of scattering in scalar field theory I
[Day 1, Lect 2] Quantum simulation of scattering in scalar field theory I
[Day 2, Lect 1] Hamiltonian formulations of gauge theories I (Abelian)
[Day 2, Lect 2] Hamiltonian formulations of gauge theories II (Non-Abelian)
[Day 3, Lect 1] Time evolution in gauge theories I (Abelian examples)
[Day 3, Lect 2] Time evolution in gauge theories II (Non-Abelian examples)
[Day 4, Lect 1] Vacuum, hadronic, and thermal state preparation in gauge theories I
[Day 4, Lect 2] Vacuum, hadronic, and thermal state preparation in gauge theories II
[Day 5, Lect 1] Physics targets: scattering and decay amplitudes, structure functions, transport coefficients, (dynamical) phases and phase transitions, thermalization
[Day 5, Lect 2] Entanglement in quantum field theories
~
\subsection{First week}

\begin{table}
\caption{


\begin{itemize}
\item
  Definitions, Linear Algebra reminder, Hilbert Space, Operators on
  Hilbert Spaces, Composite Systems

  \begin{itemize}
  \item
    Definitions
  \item
    Mathematical notation, Hilbert spaces and operators
  \item
    Computational basis
  \item
    Description of Quantum Systems and one-qubit systems
  \item
    States in Hilbert Space, pure and mixed states
  \end{itemize}
\end{itemize}

    \subsubsection{Week 1, Tuesday}\label{week-1-tuesday}

\begin{itemize}
\item
  Composite Systems and Tensor Products

  \begin{itemize}
  \item
    Spectral decomposition and measurements
  \item
    Density matrices
  \item
    Simple Hamiltonians and other operators
  \end{itemize}
\end{itemize}

    \subsubsection{Week 1, Wednesday}\label{week-1-wednesday}

\begin{itemize}
\item
  Density matrices and Measurements

  \begin{itemize}
  \item
    Spectral decomposition and measurements
  \item
    Density matrices
  \item
    Entanglement and Schmidt decomposition
  \item
    Entropies
  \end{itemize}
\end{itemize}

    \subsubsection{Week 1, Thursday}\label{week-1-thursday}

\begin{itemize}
\item
  Entanglement and entropies

  \begin{itemize}
  \item
    Entanglement and Schmidt decomposition
  \item
    Entropy as a measurement of entanglement
  \item
    Introduction to gates and calculations
  \end{itemize}
\end{itemize}

    \subsubsection{Week 1, Friday}\label{week-1-friday}

\begin{itemize}
\item
  Starting with project work

  \begin{itemize}
  \item
    Coding examples, with applications to Bell states and simple
    systems
  \item
    How to set up gates and circuits.
  \item
    Writing own code and comparing with existing computational
    frameworks
  \end{itemize}
\end{itemize}


\subsection{Tentative schedule for second week}


    \subsubsection{Week 2, Monday}\label{week-2-monday}

\begin{itemize}
\item
  Quantum gates and circuits and Hamiltonians
    \textcolor{orange}{RL: This day could be just ``Quantum algorithms'' and include some subset of DJ, BV, Simon, Shor, Grover, etc. There should be some discussion of quantum advantage via Shor, random circuit sampling, and/or boson sampling. VQE could be after this.}

  \begin{itemize}
  \item
    Quantum gates and operations and simple quantum algorithms
  \item
    Discussion of the Varational Quantum Eigensolver (VQE) algorithm and
    discussions of computational project
  \end{itemize}
\end{itemize}

    \subsubsection{Week 2, Tuesday}\label{week-2-tuesday}

\begin{itemize}
\item
  Algorithms for solving quantum mechanical problems

  \begin{itemize}
  \item
    Variational Quantum Eigensolver and discussion of codes\ale{This seems repeated from the previous day, how about we dedicate the day to Hamiltonian dynamics and introduce Trotter-Suzuki here}
  \item
    Simulation of Hamiltonian dynamics, focus on simple Hamiltonians \ale{Heisenberg chain?}
  \end{itemize}
\end{itemize}

    \subsubsection{Week 2, Wednesday}\label{week-2-wednesday}

\begin{itemize}
\item
  Solving quantum mechanical problems

  \begin{itemize}
  \item
    Introducing the Lipkin model
  \item
    Analyzing the solution of eigenvalue problems and the VQE using the
    Lipkin model. Discussion of ansatzes for initializations of qubits
  \item
    Comparison with the nuclear shell model
  \end{itemize}
\end{itemize}

    \subsubsection{Week 2, Thursday}\label{week-2-thursday}

\begin{itemize}
\item
  Second quantization and Hamiltonians for quantum computing

  \begin{itemize}
  \item
    Other Hamiltonians (Pairing \ale{Since we are doing nuclear physics how about pionless-EFT instead?}) and the Jordan-Wigner transformation
  \item
    Suzuki-Trotter expansion \ale{This should be moved to Tuesday of week 2. Given the scope of the day I would maybe do some more advanced material on fermions like e.g. fermionic swap-networks and alternative encoding schemes}
  \end{itemize}
\end{itemize}

    \subsubsection{Week 2, Friday}\label{week-2-friday}

\begin{itemize}
%\tightlist
\item
  Rodeo algorithm and comparison with the VQE algorithm
\end{itemize}


\subsection{Tentative schedule third  week}


    \subsubsection{Week 3, Monday}\label{week-3-monday}

\begin{itemize}
\item
  Do we need this? Quantum Fourier transforms and fast Fourier transform\ale{How about we focus this day instead on more advanced simulation techniques introducing things like Linear Combination of Unitaries (LCU), Block encodings and maybe Quantum Signal Processing (QSP). I feel like we should also dedicate a bit of time to discuss Quantum Phase Estimation since that's something that needs to be covered in an introductory course of QC, not sure where to put it though. If needed we could use the bulk of this day for that}
  \begin{itemize}
  \item
    Do we need this? Implementing Quantum Fourier transforms
  \item
    Other algorithms
  \end{itemize}
\end{itemize}

    \subsubsection{Week 3, Tuesday}\label{week-3-tuesday}

\begin{itemize}
%\tightlist
\item
  Quantum computing for field theory
\end{itemize}

    \subsubsection{Week 3, Wednesday}\label{week-3-wednesday}

\begin{itemize}
%\tightlist
\item
  Quantum computing for field theory
\end{itemize}

    \subsubsection{Week 3, Thursday}\label{week-3-thursday}

\begin{itemize}
%\tightlist
\item
  Quantum error correction\ale{(I feel like we should also say something about error mitigation but I am not sure how to fit that in the current schedule. Maybe we can compress a bit the material of the first week to make room? If we are targeting PhD students in physics we can probably cover those things  bit more quickly. Otherwise we can switch quantum error correction with quantum error mitigation)}
  \ale{\begin{itemize}
      \item Kraus representation and error channels
      \item standard [9,1,3] code and threshold theorem
  \end{itemize}}
\end{itemize}

    \subsubsection{Week 3, Friday}\label{week-3-friday}

\begin{itemize}
%\tightlist
\item
  Summary of course
\end{itemize}



\begin{document}


\end{document}
